Handling missing data in smartphone location logs

### 2017-12-01 thesisReportApa.tex
made it APAstyle

                      Correspondence concerning this article should be addressed to Boaz Sobrado, Postal address. E-mail: \href{mailto:boaz@boazsobrado.com}{\nolinkurl{boaz@boazsobrado.com}}
                }

  \abstract{Using objective location data to infer the mobility measures of
individuals is highly desirable, but methodologically difficult. Using
commercially gathered location logs from smartphones holds great
promise, as they have already been gathered, often span years and can be
associated to individuals. However, due to technical constraints this
data is more sparse and inaccurate than that produced by specialised
equipment. In this paper we present a model which leverages the
periodicity of human mobility in order to impute missing data values.
Moreover, we will assess the performance of the model relative to
currently used methods, such as linear interpolation.}
  \keywords{GPS, Human Mobility \\

    \indent Word count: X
  }

@@

How active people are and how they interact with their environment
affects a wide range of outcomes including health, income and social
capital (Goodchild \& Janelle, 2010). A better understanding of both
within-person and between-person variability in geospatial patterns
could be conducive to better social, health and urban-planning policies.
Yet a large part of studies on human mobility are largely based on
pen-and-paper travel diaries. These surveys have known methodological
flaws, such as the short period of data collection (due to costs and
burden to respondents), the underreporting of short trips (Wolf,
Oliveira, \& Thompson, 2003) and the underestimation of the duration of
commutes (Delclòs-Alió, Marquet, \& Miralles-Guasch, 2017).

These obstacles can be overcome by using objective data on human
mobility. The Global Positioning System (GPS), which uses the distance
between a device and a number of satellites to determine location,
provides such data. Within behavioural science, this type of data has
been used to investigate topics such as the effects of the food
environment on eating patterns (Zenk, Schulz, \& Odoms-Young, 2009), the
movement correlates of personality (G. M. Harari et al., 2016), academic
performance (Wang, Harari, Hao, Zhou, \& Campbell, 2015) and bipolar
disorder (Palmius et al., 2017).

In most of these studies participants are given a specialised device,
resulting in accurate mobility GPS data (\emph{specialised logs}).
However, Barnett and Onnela (2016) point out that these studies are not
scalable due to cost and burden to participants. Moreover, they may be
biased because of the introduction of a new device to the participant's
life. Because of these drawbacks, specialised logs usually span a short
amount of time. Barnett and Onnela (2016) advocate installing a
custom-made tracking app on user's phones (\emph{custom logs}). Another
solution is to take advantage of existing smartphone location logs, such
as Google Location History, which store location information of millions
of users spanning several years (Location History, 2017)
(\emph{secondary logs}). By law, logs can be accessed and shared by
users for free (Commission, 2017). Yet, because they were created for
non-academic purposes under engineering constraints, the sensors do not
monitor continuously and the resulting logs can be sparse and
inaccurate. Hence, two important challenges are dealing with measurement
noise and missing data.

Missing data is a pervasive issue as it can arise due to multiple
reasons. Technical reasons include signal loss, battery failure and
device failure. Behavioural reasons include leaving the phone at home,
switching the phone off, switching location measurements off, and so on.
As a result, applied researchers are often left with wide temporal gaps
with no measurements. For instance, different groups studying the effect
of bipolar disorder on human movement have reported missing data rates
between 30\% to 50\% (Grünerbl et al., 2015; Palmius et al., 2017; Saeb
et al., 2015). Similar trends are consistently reported in other fields
(e.g. G. M. Harari et al., 2016; Jankowska, Schipperijn, \& Kerr, 2015).

There is currently no golden standard in how to deal with missing data
in custom or secondary logs (Barnett \& Onnela, 2016). Traditional
missing data methods, such as mean imputation, cannot be used easily in
spatiotemporal data because the measurements are correlated in time and
space. For example, assume the simplistic case that an individual spends
almost half her time at work, half her time at home and a small amount
of time commuting between the two along a angled path. Using mean
imputation would result in imputed values of her being at the geometric
midpoint between the home and work for all missing values, even though
she has never been there and never will be. Worringly, Jankowska et al.
(2015) have pointed out that there is often little transparency
regarding decisions of how to deal with missing data.

The accuracy of GPS measurements in smartphones is substantially lower
than in professional grade GPS trackers. For example, Android phones
collect location information through a variety of methods, such as from
WiFi access points, cellphone triangulation, and GPS measurements. They
use different methods due to computational and battery constraints (M.
Y. Chen et al., 2006; LaMarca et al., 2005).In professional grade GPS
trackers less than 80\% of measurements fall within 10 meters of the
true location. GPS measures are reported to be most inaccurate in high
density urban locations and indoors (S. Duncan et al., 2013; Schipperijn
et al., 2014). Unfortunately for social scientists, this happens to be
where most people in the developed world tend to spend most of their
time.

On the other hand, noisy data can lead to inaccurate conclusions if it
is not accounted for, such as overestimating the movement of
individuals. For instance, suppose that a naive researcher calculates
the distance travelled by an individual by drawing a line between each
measured point and calculating the sum of the length of all of these
lines. If there is noise, the measurements will vary even though the
individual is not moving. If the measurements are frequent, then the
researcher will end up with a lot of movement, even though the
individual did not move at all. The problem is further complicated by
the fact that missing data and noisy measurements are related to each
other. For instance, methods used by researchers to reduce noise, such
as throwing out inaccurate measurements (e.g. Palmius et al., 2017) can
exacerbate the severity of the missing data problem.

In this paper we will compare methods used to deal with measurement
error and missing data in mobility patterns from secondary GPS logs.

\section{A concrete example}\label{a-concrete-example}

Given that there is little literature on dealing with missing data in
custom or secondary logs it is worth illustrating the typical
characteristics of this data using an example data set. The example
dataset comes from the Google Location History of a single individual
and spans from January 2013 to January 2017. It was recorded with
multiple different Android devices and contains 814 941 measurements,
with approximately 742 measurements per day
(\(\widehat{\sigma}\)=868.15). The dataset contains a wide range of
variables including inferred activity and velocity. For the purposes of
this paper we will focus only on latitude, longitude, accuracy (defined
below) and a timestamp.

\subsection{Location logs and
notation}\label{location-logs-and-notation}

GPS measurements report our location on a three dimensional planet, yet
we are interested in placing these measurements on a two dimensional
map. Projecting three dimensional measurements onto a two dimensional
plane results in errors, in order to minimise these errors we borrow an
error minimising projection method from Barnett and Onnela (2016).

Let a persons' true location on this two-dimensional plane be
\(G(t) = [G_x(t) G_y(t)]\) where \(G_x(t)\) and \(G_y(t)\) denote the
location of the individual at time \(t\) on the x-axis and y-axis
respectively. Moreover, let \(D \in \mathbb{R}^2\) be the recorded data
containing lattitude and longitude. In addition, let \(a\) denote the
estimated accuracy of the recorded data. accuracy. \(G(t)\), \(D\) and
\(a\) are indexed by time labled by the set \(T = t_1 < ... < t_{n+1}\).
For simplicity, let each entry in the discrete index set \(T\) represent
a 5 minute window. The measure of accuracy \(a_t\) is given in meters
such that it represents the radius of a 67\% confidence circle. If
\(D_t = \emptyset\) it is considered \emph{missing} and it is not
missing otherwise.

@@

In the example data set the distribution of \(a\) is highly right
skewed, with a median of 28, \(\mu = 127\) and the maximum value at 26
km. Palmius et al. (2017) note that in their Android based custom logs
inaccurate location values are interspersed between more accurate
location values at higher sample rates per hour. We observe similar
patterns in secondary logs. Figure 1 shows how accuracy tends to vary as
a function of user behaviour, time and location. There are several
recurring low-accuracy points, most likely cell-phone triangulation
towers.

\begin{figure}
\includegraphics[width=1\linewidth]{img/journeyTillMiddayBoaz} \caption{Measurement accuracy of each logged measurement in a morning journey. The red circles denote the accuracy of all logged measurement points (the raw data). The points connected in time are connected by a line. The blue line shows the path without the most inaccurate (accuracy > 400 meters) points filtered out. The red line shows the path with all measurements included. }\label{fig:accuracyPlot}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{img/accuracyLocShift} \caption{The accuracy of raw measures with time on the x-axis and accuracy on the y-axis. The colour scale shows the distance between the measurement and the previous point.}\label{fig:accuracyPlot2}
\end{figure}

@@

Over 54\% of the data is missing for the entire duration of the log.
However,this is misleading as there are several long periods with no
measurements whatsoever. The structure of missingness of a day with
measurements is shown below.

\begin{figure}
\includegraphics[width=1\linewidth]{img/missingBoaz5minExample} \caption{Example of missing data over a day. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. For this day there were several periods with no information. These points are filled with red and lie on the x-axis.}\label{fig:measurementsPerDay}
\end{figure}

\subsection{Current methods for imputing missing spatiotemporal
data}\label{current-methods-for-imputing-missing-spatiotemporal-data}

As we have mentioned before, Jankowska et al. (2015) point out that
there is little transparency regarding decisions of how to deal with
missing data in GPS mobility logs. We suspect this is because the highly
interdisciplinary nature of the problem means researchers are unaware of
potential solutions. For this reason, we consider it important to
briefly discuss a few methods one could consider in order to deal with
measurement inaccuracy and missing data problems in spatiotemporal data.
In doing so, we will argue that state space models (SSMs) and
spatiotemporal imputation methods, which are extensively used to model
missing and noisy spatiotemporal data, are not well suited to deal with
human mobility patterns. Moreover, we discuss in detail two approaches
by Palmius et al. (2017) and Barnett and Onnela (2016) which deal
explicitly with missing data in mobility patterns from smartphone GPS
logs.

There is a vast literature of using SSMs to improve measurements
accuracy and deal with missing data. Behavioural ecologists for
instance, have used SSMs extensively to explain how animals interact
with their environment (Patterson, Thomas, Wilcox, Ovaskainen, \&
Matthiopoulos, 2008). These models can be quite complex, for example
Preisler, Ager, Johnson, and Kie (2004) uses Markovian movement
processes to characterise the effect of roads, food patches and streams
on cyclical elk movements. The most well studied SSM is the Kalman
filter, which is the optimal algorithm for inferring linear Gaussian
systems. In fact, the extended Kalman filter is the de facto standard
for GPS navigation (Z. Chen \& Brown, 2013). The advantage of state
space models is that they are flexible, deal with measurement
inaccuracy, include information from different sources and can be used
in real time.

However, the main limitation of SSMs is that they ignore the fact that
humans have regular movement routines, such as going to work or shopping
for groceries on weekends. This limitation is due to the fact that SSMs
are based on the Markov property. Thus, the estimated location \(G(t)\)
at timepoint \(t\) is often based only upon measurements \(D_t\),
\(D_{t-1}\) and ignores all \(D_{t-i}|i\geq2\). Hierarchical structuring
and conditioning on a larger context have been suggested as ways to
improve the performance of Markovian models, but these solutions are
often computationally intractable or unfeasible (Sadilek \& Krumm,
2016). For this reason we do not consider SSMs to be useful for imputing
missing data, although they could be of use in filtering noise.

In addition to SSMs there are spatiotemporal imputation methods often
used in climate or geological research for estimating missing data. For
instance, Feng, Nowak, O'Neill, and Welsh (2014) illustrate their CUTOFF
method, which relies on estimating missing values using the nearest
observed neighbours in time, using rainfall data from dozens of gauging
stations across Australia. Similarly, Z. Zhang et al. (2017) use a
variety of machine learning methods to present their model based on
underground water data in China.

The difficulty of applying this class of spatiotemporal imputation
models to human mobility tracks lies in the fact that these models
generally assume fixed measurement stations (such as rainfall gauging
stations). While Feng et al. (2014) claim their model could be used to
establish mobility patterns, ostensibly by dividing the sample space
into rasters analogous to measurement stations indicating a probability
of the individual being there, this seems to be computationally
unfeasible. To our knowledge such models have not been implemented for
mobility traces.

@@

\subsection{Methods}\label{methods}

@@

The data used to train the imputation methods was collected between 2013
and 2017 on different Android devices from several individuals (table
1).

\begin{table}

\caption{\label{tab:datadetailsTable}Table with descriptives about the data sets used to build the imputation methods. }
\centering
\begin{tabular}[t]{l|r|r|r|r|r}
\hline
Log duration & Logged days & Observations & Missing days & Missing data & Mean Accuracy\\
\hline
From 2013-02-06 to 2017-03-29 & 1512 & 646376 & 635 & 0.22 & 127.78\\
\hline
From 2016-07-14 to 2017-05-10 & 300 & 158382 & 3 & 0.41 & 1394.60\\
\hline
From 2014-01-22 to 2017-01-23 & 1097 & 814941 & 80 & 0.25 & 121.83\\
\hline
\end{tabular}
\end{table}

@@

Analyses were performed using R and a multitude of other statistical
packages ({\textbf{???}}; Bivand, Pebesma, \& Gomez-Rubio, 2013; E. J.
Pebesma \& Bivand, 2005; R Core Team, 2017; Wickham, 2009; Wickham \&
Francois, 2016).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Mean imputation as described by Palmius et al. (2017).
\item
  The model developed by Barnett and Onnela (2016) using spatial,
  temporal and circadian proximity.
\item
  Simple linear interpolation was used as a benchmark model.
\end{enumerate}

>>>>

                      Correspondence concerning this article should be addressed to Boaz Sobrado. E-mail: \href{mailto:boaz@boazsobrado.com}{\nolinkurl{boaz@boazsobrado.com}}
                }

  \abstract{Using objective location data to infer the mobility measures of
individuals is highly desirable, but methodologically difficult. Using
commercially gathered location logs from smartphones holds great
promise, as they have already been gathered, often span years and can be
associated to individuals. However, due to technical constraints this
data is more sparse and inaccurate than that produced by specialised
equipment. In this paper we present a model which leverages the
periodicity of human mobility in order to impute missing data values.
Moreover, we will assess the performance of the model relative to
currently used methods, such as linear interpolation.}
  \keywords{Missing data, Measurement Bias,GPS, Human Mobility \\

    \indent Word count: 2205
  }

@@

How people move about in their environment affects a wide range of
outcomes, such as health, income and social capital (Goodchild \&
Janelle, 2010). A better understanding of mobility could lead to better
health and urban-planning policies. Yet a large part of studies on human
mobility are conducted with pen-and-paper travel diaries, despite well
known methodological flaws. The high cost and burden to respondents
limits the span of data collection. Short trips are frequently
under-reported (Wolf, Oliveira, \& Thompson, 2003) . The self-reported
duration of commutes is often underestimated (Delclòs-Alió, Marquet, \&
Miralles-Guasch, 2017).

These obstacles can be overcome by using objective data on human
mobility. Such data can now be obtained using the Global Positioning
System (GPS). GPS uses the distance between a device and several
satellites to determine location. GPS measurements can be used to infer
a vast range of socioeconomic and behavioural measures, including where
the individual lives, how much time he or she spends at home and where
(and how) they travel.

Within behavioural sciences, researchers have used GPS data to
investigate wideranging topics. For example, Zenk, Schulz, and
Odoms-Young (2009) investigate the effects of the food environment on
eating patterns. G. M. Harari et al. (2016) look at the movement
correlates of personality, finding that extrovers individuals spend less
time at home than introverts. Wang, Harari, Hao, Zhou, and Campbell
(2015) looks at how academic performance is affected by movement
patterns. Palmius et al. (2017) use mobility patterns to predict bipolar
depression.

In most studies participants receive a specialised GPS devices to track
their movement. We call resulting logs \emph{specialised logs}. Barnett
and Onnela (2016) point out several methodological issues with these
studies. Like studies using pen-and-paper travel diaries, collecting
specialised logs is costly and places a high burden on participants.
Besides, introducing a new device to the participant's life may bias
their behaviour. Due to these drawbacks specialised logs usually span a
short amount of time. Barnett and Onnela (2016) advocate installing a
custom-made tracking app on user's phones (\emph{custom logs}). Another
solution is to take advantage of existing smartphone location logs
(\emph{secondary logs}) . For instance, Google Location History contains
information on millions of users (Location History, 2017). Often,
secondary logs span several years. By law, secondary logs are accessible
to users for free (Commission, 2017). Yet, secondary logs also present
methodological challenges. They were created for non-academic purposes
under engineering constraints (detailed subsequently). These constraints
mean that sensors do not track users continuously, meaning that the
resulting logs can be sparse and inaccurate. Hence, two important
challenges are dealing with measurement noise and missing data.

Missing data is a pervasive issue as it can arise due to several
reasons. Technical reasons include signal loss, battery failure and
device failure. Behavioural reasons include leaving the phone at home or
switching the device off. As a result, secondary logs often contain wide
temporal gaps with no measurements. For instance, several research
groups studying mental health report missing data rates between 30\% to
50\% (Grünerbl et al., 2015; Palmius et al., 2017; Saeb et al., 2015).
Other researchers report similar trends in different fields (e.g. G. M.
Harari et al., 2016; Jankowska, Schipperijn, \& Kerr, 2015).

There is no golden standard for dealing with missing data in GPS logs
(Barnett \& Onnela, 2016). Importantly, spatiotemporal data measurements
are often correlated in time and space. This means that common methods,
such as mean imputation, are unsuitable. For example, imagine an
individual who splits almost all her time between work and home. Suppose
she spends a small amount of time commuting between the two along a
circular path. Using mean imputation to estimate her missing
coordinates, we impute her to be at the midpoint between home and work.
She has never and will never be there! Worryingly, there is little
transparency on how researchers deal with missing data (Jankowska et
al., 2015).

The accuracy of smartphone location measurements is substantially lower
than that of professional GPS trackers. Android phones collect location
information through a variety of methods. Other than GPS measurements,
Androids use less-accurate heuristics such as WiFi access points and
cellphone triangulation. Different methods are used because of
computational and battery constraints. GPS is the most energy consuming
sensor on most smartphones (M. Y. Chen et al., 2006; LaMarca et al.,
2005). In professional GPS trackers less than 80\% of measurements fall
within 10 meters of the true location. GPS measures are most inaccurate
in dense urban locations and indoors (S. Duncan et al., 2013;
Schipperijn et al., 2014). Unfortunately for researchers, this is where
people in the developed world spend most of their time.

Noisy data can lead to inaccurate conclusions if it is not accounted
for. Suppose we wish to calculate an individual's movement in a day. A
simple approach would be to calculate the sum of the distance between
each measurement. But if there is noise, the coordinates will vary even
though the individual is not moving. If the measurements are frequent
and noisy, we will calculate a lot of movement, even if the individual
did not move at all! This issue is also visualised in Figure
\ref{fig:accuracyPlot}. The problem is further complicated because
missing data and noisy measurements are related. Methods used by
researchers to reduce noise, such as throwing out inaccurate
measurements (e.g. Palmius et al., 2017), can exacerbate the severity of
the missing data problem.

In this paper we will explore in detail the problem of missing data and
measurement error in secondary location logs. Moreover, we will compare
methods used to deal with these problems.

\subsection{A concrete example}\label{a-concrete-example}

There is little literature on dealing with missing data in custom or
secondary logs. Thus it is worth illustrating the typical
characteristics using an example data set. The example data set comes
from the Google Location History of a single individual. It spans from
January 2013 to January 2017 and contains 814 941 measurements. The data
set contains a multitude of variables, including inferred activity and
velocity. We will focus on measurements of latitude, longitude, accuracy
(defined below). All measurements are paired with a timestamp.

\subsubsection{Aggregate measures}\label{aggregate-measures}

Social scientists are most interested in aggregating spatiotemporal data
to more socially relevant information, such as distance travelled. As we
discussed earlier, aggregations without data processing can be highly
biased. However, as an example we calculate time spent at home of the
user for each day of the week in the month of February (Figure
\ref{fig:aggrePlot}). Between individuals, time spent at home has been
found to be a reliable predictor of extraversion (G. M. Harari et al.,
2016).

\begin{figure}
\includegraphics[width=1\linewidth]{img/timeUse} \caption{Proportion of time spent at home in February 2017. We estimate this by calculating the mean lattitude and longitude for every 5 minute time period in the month. Missing values are filled in by the previous observed value. Behavioural trends are evident, the user spends more time at home on weekends than on weekdays. Moreover, the variance is greater on weekends, due to travel.}\label{fig:aggrePlot}
\end{figure}

@@

Google location history provides a measure of accuracy that is given in
meters such that it represents the radius of a 67\% confidence circle.
In the example data set the distribution of accuracy is highly right
skewed, with a median of 28, \(\mu = 127\) and the maximum value at 26
km. Palmius et al. (2017) note that in their Android based custom logs
inaccurate location values are interspersed between more accurate
location values at higher sample rates per hour. We observe similar
patterns in secondary logs. Figure \ref{fig:accuracyPlot} shows how
accuracy can vary as a function of user behaviour, time and location.
Inaccurate measures are often followed by more accurate measures. Most
notably, low accuracy often (but not always) is associated with movement
(Figure \ref{fig:accuracyPlot2}). Stationary accuracy varies depending
on phone battery level, wifi connection and user phone use. There are
several recurring low-accuracy points, possibly the result of cell-phone
tower triangulation.

\begin{figure}
\includegraphics[width=1\linewidth]{img/journeyTillMiddayBoaz} \caption{Measurement accuracy of each logged measurement of a morning journey on February 15th 2017. This includes all measurements from midnight to midday. The red circles denote the accuracy of all logged measurement points (the raw data). The points connected in time are connected by a line. The blue line shows the path without the most inaccurate (accuracy > 400 meters) points filtered out. The red line shows the path with all measurements included. }\label{fig:accuracyPlot}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{img/accuracyLocShift} \caption{Measures of user activity and measurement accuracy on February 15th 2017.The upper chart shows the distance from the next measured point in meters over the course of the day. All journeys are marked with a red line. The first peak corresponds to the first journey from the user's home to a gym around 8am. The second, smaller peak before 10 reflects a journey from the gym to the nearby lecture theatre. Both journeys can be seen in Figure 1. All other journeys are not shown in Figure 1. The large jump between journey 5 and 6 is measurement error. The lower chart shows the accuracy over the course of the day. The figure shows that measurement inaccuracy is sometimes related to the movement of the individual.}\label{fig:accuracyPlot2}
\end{figure}

@@

Over 54\% of the data is missing for the entire duration of the log.
This may be misleading as there are several long periods with no
measurements whatsoever (see Figure \ref{fig:longMeasurementsPerDay}).
For days which were not entirely missing, approximately 22\% of all five
minute segments were missing. The structure of missingness of a day with
measurements is shown in Figure \ref{fig:measurementsPerDay}. As you can
see, there are several long periods over the course of the log for which
there are no measurements. Moreover, even during a single day there are
continuous periods where there is missing data, mostly during the late
hours of the night in this case.

\begin{figure}
\includegraphics[width=1\linewidth]{img/missingdayBoaz5min} \caption{Example of missing data over the entire duration of the log. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. For this day there were several periods with no information. These points are filled with red and lie on the x-axis.}\label{fig:longMeasurementsPerDay}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{img/missingBoaz5minExample} \caption{Example of missing data on February 15th 2017. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. For this day there were several periods with no information. These points are filled with red and lie on the x-axis.}\label{fig:measurementsPerDay}
\end{figure}

\section{Methods}\label{methods}

\subsection{Two-dimensional projections and
notation}\label{two-dimensional-projections-and-notation}

GPS measurements provide us with coordinates on the surface of the
earth. Because most mobility metrics are computed for data in
\(\mathbb{R}^2\), we are interested in mapping these \(\mathbb{R}^3\)
measurements on a 2D Euclidean plane. Projecting three dimensional
measurements onto a two dimensional plane results in distortion. To
minimise errors we borrow an error minimising projection method from
Barnett and Onnela (2016).

Having thus converted lattitude and longitude onto coordinates unique to
each individual, let a person's true location on this two-dimensional
plane be \(G(t) = [G_x(t) G_y(t)]\) where \(G_x(t)\) and \(G_y(t)\)
denote the location of the individual at time \(t\) on the x-axis and
y-axis respectively. Moreover, let \(D \in \mathbb{R}^2\) be the
recorded data containing lattitude and longitude. In addition, let \(a\)
denote the estimated accuracy of the recorded data. accuracy. \(G(t)\),
\(D\) and \(a\) are indexed by time labled by the countable set
\(t = t_1 < ... < t_{n+1}\). For simplicity, let each entry in the
discrete index set \(t\) represent a 5 minute window. The measure of
accuracy \(a_t\) is given in meters such that it represents the radius
of a 67\% confidence circle. If \(D_t = \emptyset\) it is considered
\emph{missing} and it is not missing otherwise.

When several data sets are available from individuals living in
overlapping areas we can construct a \(t \times i\) matrix \(M\) where
the entry \(M(t,i)\) contains \(G(t)\) for the individual \(i\).

\subsection{Selecting candidate
models}\label{selecting-candidate-models}

There is no golden standard or established practice in how to deal with
missing data in GPS logs. Researchers are generally vague about what
practices they follow (Jankowska et al., 2015). Ostensibly this is
because they are unaware of possible solutions. In an attempt to
elucidate the topic, we explore potential solutions. We will argue that
extensively used spatiotemporal methods, such as state space models
(SSMs), are not well suited to deal with human mobility patterns. We
also discuss in detail two approaches which deal explicitly with
mobility patterns from custom or secondary logs (Barnett \& Onnela,
2016; Palmius et al., 2017 ).

There is a vast literature on using SSMs in spatiotemporal statistics.
For example, ecologists have used SSMs to explain how animals interact
with their environment (Patterson, Thomas, Wilcox, Ovaskainen, \&
Matthiopoulos, 2008). These models can be quite complex. Preisler, Ager,
Johnson, and Kie (2004) uses Markovian movement processes to
characterise the effect of roads, food patches and streams on cyclical
elk movements. The most well studied SSM is the Kalman filter, which is
the optimal algorithm for inferring linear Gaussian systems. The
extended Kalman filter is the de facto standard for GPS navigation (Z.
Chen \& Brown, 2013). The advantage of state space models is that they
are flexible, deal with measurement inaccuracy, include information from
different sources and can be used in real time.

For us, the main limitation of SSMs is that they ignore regular movement
routines. For instance, humans tend to go to work on weekdays and sleep
at night. Because SSMs are based on the Markov property, they cannot
incorporate this information. The estimated location \(G(t)\) at
timepoint \(t\) is often based only upon measurements \(D_t\),
\(D_{t-1}\) and ignores all \(D_{t-i}|i\geq2\). Hierarchical structuring
and conditioning on a larger context have been suggested as ways to add
periodicity to Markovian models. These solutions are often
computationally intractable or unfeasible (Sadilek \& Krumm, 2016). For
this reason we do not consider SSMs to be useful for imputing missing
data. Nonetheless, they could be of use in filtering noise.

In climate or geological research spatiotemporal imputation methods are
often used. For instance, the CUTOFF method estimates missing values
using the nearest observed neighbours (Feng, Nowak, O'Neill, \& Welsh,
2014 ). The authors illustrate their example using rainfall data from
gauging stations across Australia. Similarly, Z. Zhang et al. (2017) use
a variety of machine learning methods to impute missing values. The
example provided relates to underground water data. Generally these
models assume fixed measurement stations (such as rainfall gauging
stations).

For this reason they cannot be easily applied to missing mobility
tracks. Feng et al. (2014) claim their model could be used to establish
mobility patterns. This may be possible by dividing the sample space
into rasters. Each raster would be analogous to a measurement station.
These artificial stations could \enquote{measure} the probability of the
individual being there. To our knowledge such models have not been
implemented for mobility traces and seems computationally inefficient.

@@

@@

The data used to train the imputation methods was collected between 2013
and 2017 on different Android devices from several individuals (Table
\ref{tab:datadetailsTable}).

\begin{lltable}
\begin{longtable}{llllll}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
\caption{\label{tab:datadetailsTable}Table with descriptives about the data sets used to build the imputation methods. Missing data stands for the proportion of missing 5 minute windows within days that were not missing entirely.}\\
\toprule
Log duration & Logged days & Observations & Missing
days & Missing
data & Mean Accuracy\\
\midrule
2013-02-06 to 2017-03-29 & 1,512.00 & 646,376.00 & 635.00 & 0.22 & 127.78\\
2016-07-14 to 2017-05-10 & 300.00 & 158,382.00 & 3.00 & 0.41 & 1,394.60\\
2014-01-22 to 2017-01-23 & 1,097.00 & 814,941.00 & 80.00 & 0.25 & 121.83\\
\bottomrule
\end{longtable}
\end{lltable}

@@

Analyses were performed using R and a multitude of other statistical
packages (Bivand, Pebesma, \& Gomez-Rubio, 2013; E. J. Pebesma \&
Bivand, 2005; R Core Team, 2017; Wickham, 2009; Wickham \& Francois,
2016).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Mean imputation as described by Palmius et al. (2017).
\item
  The model developed by Barnett and Onnela (2016) using both spatial
  and temporal proximity.
\item
  Simple linear interpolation was used as a benchmark model.
\end{enumerate}
